Towards typed repositories of proofs \\[0.6em] \small \textsf{MIPS 2010}

### 2011-03-13 doc/talks/MIPS2010/handout.tex
Moved coq/ papers/ talks/ to doc/. Removed old implementation in experiments/
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@341 403e655e-980a-47bd-8c60-1ee5348bb6b7

I am going to present a work-in-progress, realized together with Yann
Régis-Gianas. It is far from a finished product, so feel free to
interrupt, comment or criticize. So this talk will be about some
observations and directions to improve the way we edit and store
formal proofs in proof assistants.

@@

Let me start with a simple observation relating the activity of both
the working mathematician and the computer scientist. Both activities
could be percieved, from the outside, as apodictic because the final
object of their study, the proof or the program, seems to be
unquestionably valid (or not).

If we observe the day-to-day work of these scientists though, their
workflow resembles a lot more that of experimental sciences: it is
highly non-linear, involving multiple parallel experiments, fixes,
backtrack on previous modifications\ldots\ eventually validated or
invalidated by some \emph{a posteriori} criterion, the absence of
bugs, the validity of a proof.

It is the case for programs, for classical, paper mathematics so it
should be the case for formal mathematics, edited in a proof
assistant.

Since the advent of modern proof assistants, large bodies of
mathematics have been formalized and the way people managed these
developments is very similar to the one used in software development
for years: we write proofs in text files, which are linear, and launch
proof-checking when we are done. To avoid re-checking the whole
development, we split it into inter-dependent files and re-check only
those who have changed. To keep track of old versions, we use
line-based version control systems that have no awareness of the
actual content of the files. In a sense, the non-linearity of proof
development is managed outside the scope of the tool, by manual,
textual transformations that are often very unsatisfactory.

The paradox about this situation is that these very tools (Coq,
Matita, Agda) can be used to reason about formal languages and
especially proof languages, some of these tools are even dedicated to
the mechanization of the metatheory of proof languages. But we still
use this legacy generic tools to manage our developments.

@@

Instead of writing our complete development in one unique file and let
the program check it when done, we can split it in several files or
modules and establish dependencies among them: only those who changed
(and their dependencies) have to be rechecked.

But this separation serves two different purpose, and the fact that
they coincide is pure contingency: the purpose of splitting a
development into logical units, and the purpose of providing a coarse
notion of dependency to ease the non-linear process of proof editing
by accelerating recompilation upon changes.

Secondly, the traditional interaction loop makes unnecessary the
recompilation upon local changes by providing a global undo feature of
the whole state of the proof assistant. But if I want to change this
proof, I have to backtrack all the way here (because the green region
is locked), do my changes and then replay the whole proofs in between,
even if my change had no impact on this one, even if I just added a
comment.

@@

In the following, I will present a preliminary implementation of these
ideas. It is a meta-language for incremental proof-checking, that is a
language in which one can define abstract syntaxes, logical rules and
proofs, and allows to check validity, express the incrementality of
the constructions, and their dependency. It can be viewed as a kernel
for a generic, typed version control system, and for the first
iteration of this project, we were inspired by the way \textsf{git}
stores the history of directories.

@@

\textsf{Git} is one of the most successful version control system, and
part of its success is due to the very simple, yet very powerful
storage model. It doesn't keep track of changes \emph{per se}, but of
similarities between version. Given a base directory structure,
already committed in its database, and a new version of it with, let's
say, one file changed -- \texttt{f3.txt}, it stores the new version by
linking all common subdirectory structure to the previous
version. This way, similarities are physically relalized by this kind
of ``compression''. This is possible because every node in the tree,
directories and files, have a unique name. So actually, the database
of \textsf{git} looks like this: it associates to every name, a
content (for files) or a list of names (for directories). The
invariant is that this database forms a DAG. This naming policy makes
all content adressable.

Moreover, similarities are easily tracked because every name is
carefully chosen to reflect the \emph{content} of a file or directory.
In practice, \textsf{git} uses a hash function to produce all names in
the database: files are given the hash of their content, and
directories are given the hash of the hash of all the files it
contains, recursively.

@@

>>>>

Let me start with a simple observation on the activity of both the
working mathematician and the computer scientist. If we observe the
day-to-day work of these scientists, their workflow resembles a lot
that of researchers in experimental sciences: it is highly non-linear,
involving multiple parallel experiments, fixes, backtrack on previous
modifications\ldots\ eventually validated or invalidated by some
\emph{a posteriori} criterion, the absence of bugs, the validity of a
proof.

I am going to present a work-in-progress, realized together with Yann
Régis-Gianas. It is far from a finished product, so feel free to
interrupt, comment or criticize. It will be about some observations
and directions to improve the way we edit and store formal proofs in
proof assistants.

@@

We now have very capable tools available to reason about formal
languages and especially proof languages. Some of these tools are even
dedicated to the mechanization of the metatheory of proof languages.

But still, the way people managed large development of formal proofs
is very similar to the one used in software development for years: we
write proofs in text files, which are linear, and launch
proof-checking when we are done. To avoid re-checking the whole
development, we split it into files. To keep track of old versions, we
use line-based version control systems that have no awareness of the
actual content of the files.

In a sense, the non-linearity of proof development is managed outside
the scope of the tool, by manual, textual transformations that are
often very unsatisfactory.

@@

Instead of writing our complete development in one unique file, we can
split it in several files or modules and establish dependencies among
them: only those who changed (and their dependencies) have to be
rechecked upon a change.

But this separation serves two different purpose, and the fact that
they coincide is pure contingency: the purpose of splitting a
development into logical units, and the purpose of accelerating
recompilation upon changes.

Secondly, the traditional interaction loop facilitates local changes
by providing a global undo feature of the whole state of the tool. But
if I want to change this proof, I have to backtrack all the way here
(because the green region is locked), do my changes and then replay
the whole proofs in between, even if my change had no impact on this
one, even if I just added a comment.

@@

In the following, I will present a preliminary implementation of these
ideas. It is a meta-language for incremental proof-checking, that is a
language in which one can define abstract syntaxes, logical rules and
proofs, and allows to check validity, express the incrementality of
the constructions, and their dependency. It can be viewed as a kernel
for a generic, typed version control system.

@@

As an inspiration, let's see a very simple, common technique for
storing history of directories. We won't keep track of changes
\emph{per se}, but of similarities between version. Given a base
directory structure, already committed in its database, and a new
version of it with, let's say, one file changed -- \texttt{f3.txt}, we
stores the new version by linking all common subdirectory structure to
the previous version. This way, similarities are physically relalized
by this kind of ``compression''. This is possible if we give to every
node in the tree, directories and files, a unique name. So actually,
our database looks like this: it associates to every name, a content
(for files) or a list of names (for directories). The invariant is
that this database forms a DAG. This naming policy makes all content
adressable.

Moreover, similarities are easily tracked if every name is carefully
chosen to reflect the \emph{content} of a file or directory.  In
practice, a hash function is used to produce all names in the
database: files are given the hash of their content, and directories
are given the hash of the hash of all the files it contains,
recursively.

@@

This is exactly the storage model of \textsf{git}, a popular version
control system.
